A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility

### 2019-05-09 paper.Rmd
Fixed more typos and text

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors (citation).  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

**In this paper we focus on the practical side of sharing data for the purpose of reproducibility.** Our goal is to describe various methods in which an analyst can share their data with minimal friction. We steer clear of theoretical ideas such as the FAIR data principles [@Wilkinson2016] since they still do not help a researcher share their data. We also skip the discussion around citation and credit because data citations are still poorly tracked and there is no standardized metric or a h-index equivalent for data as of this writing.

@@

The first two of these challenges have largely been solved.

@@

Providing the right amount of documentation for your data is challenging. Too much is overwhelming; too little and you don't know anything. To illustrate this, imagine two datasets. The first is contained in a folder with dozens of data files and several papers on the topic. The second contains only a single csv file. The first is too much; understanding it requires a high time cost. The second is too little; we don't know anything about what the data is or why it was collected. The challenge is to provide the right level of detail, somewhere between these two extremes.

@@

One low cost and easy way to distribute data alongside compute would be to package the datasets as a data only package or as part of something larger where the methods are being implemented. The R ecosystem has many noteworthy examples of data-only packages. One exemplar.  One exemplar is the `nycflights13` package by Hadley Wickham (https://cran.r-project.org/web/packages/nycflights13/index.html). This package makes available airline data for all flights departing New York city in 2013 in a tidy format, with distinct tables for metadata on airlines, airports, weather, and planes. The package not only provides ready to use binary data but also ships raw data (in a folder called `data-raw`) along with scripts used to clean them. The package was originally created as a way to provide example data to teach tidy data principles and serves as a great model for how one can publish a data package in R.

@@

Of the 15539 packages on Central R Archive Network (CRAN), 6278 contain
datasets either as binary data (5903 packages) or as external datasets
(766). Binary files comprise a bulk of the data shipped in the `data`
folder (68.06%) with other plain text formats such as txt, csv, dat,
json comprising less than one percent of data shipped in packages.

@@

Another common situation that researchers face is in dealing with data that fall somewhere between small (tabular data such as csvs that are in the order of bytes to megabytes to gigabytes that can be easily compressed) to large (what falls under the umbrella of `big data`). The happy medium is generally data that are too big to fit on memory of most standard machines, but can successfully fit on disk (https://ropensci.github.io/arkdb/articles/articles/noapi.html). In this case, users who do not have the support to maintain resource intensive database servers can instead rely on light serverless databases such as MonetDB or SQLite. These databases provide disk based storage and using language agnostic interfaces, a analyst can easily query these data in manageable chunks that don't overrun memory limitations. Using software packages such as arkdb (https://ropensci.github.io/arkdb/index.html) one could easily chunk large data from flat text files to these lite databases without running into memory limitations.

@@

In **Domain Specific Venue**, data can be published in journal data papers, or venues specific to the type of data. For example, in astronomy the data is hosted at the SDSS (SDSS), Genetic data can be hosted at GenBank (@GenBank), and so on.  Some other places to put data include the open science framework: https://osf.io/, Dataverse, https://dataverse.org/, and data.world, https://data.world/. Other datasets might be produced

The purpose, use, and origin of the data is an important component to consider. Data for research has a different domain compared to data collected by governments, or by journalists. Many governments or civil organisations are now making their own data available through a government website interface. Media and journalism groups are also making their data available either through places like GitHub (https://github.com/the-pudding/data), or organisations such as data.world, or may self host their data (https://data.fivethirtyeight.com/).

@@

Below is an example data dictionary table taken from the [tidy tuesday repository on incarceration trends](https://github.com/rfordatascience/tidytuesday). This includes information on the variable the class (type) of the variable, and a longer description of the variable.

@@

Other licenses or notices to be aware of are **copyrighted data**, and **data embargos**. If you are working with data **already copyrighted**, (for example under CC BY or CC0), you must give follow appropriate guidelines for giving credit. Data may also be under **embargo**. This means data cannot be shared more widely until a specific release time. If sharing data under an embargo, include detailed information on the embargo requirements in: the README, and in separate correspondence with those who receive the data.

@@

TODO: Needs a succint summary here

@@

4. **Provide a machine readable format** for your data that describes what each column contains. This can be a json schema file, which also helps websites like Google Dataset Search automatically discover your data.
5. **Provide the data in its rawest form** in a folder called "data-raw". This
 ensures provenance of your data by keeping a master copy from which all
other tidied up data can be created.
6. **Provide [open source?] scripts** used to clean data from rawest form into tidy/analysis ready format. These describe the steps taken to prepare the data,
which helps explain and document any decisions taken during the cleaning phase.
These should ideally operate on some raw data stored in the "data-raw" folder (rule 5).
7. **Keep additional copies in more accessible locations**: Even if you archive the data somewhere long-term, keep a second (or additional copies) in locations that are more readily and easily accessible for data analysis workflows. These include services like GitHub, GitHub LFS and others where fast CDNs make access and reuse practical, but fall back to Zenodo or similar in case these services were to go away.
8. **Use a hash function like MD5 checksum** to ensure that the data you deposited are the same that are being made available to others. Hash functions not only remove redundancy by removing the need to download the same data again, but also help ensure that the data are valid.
9. **Only add a citation if your data has a DOI**. A citation only makes sense when your data has a DataCite compliant DOI, which is automatically provided when data is published in repositories like Zenodo and Dryad. Even then a citation may not accrue references, but without a DOI, it is gauraunteed not to.
10. Store a copy of your data in simple formats such as csv: CSV files follow a flat and extremely simple schema and can be opened by any text editor. These files can also be easily parsed by any data science programming language. Modern software packages allow for fast reads without consuming too much memory and these tools also allow for easy modification of the data.

>>>>

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors (CITATION).  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

**In this paper we focus on the practical side of sharing data for the purpose of reproducibility.** Our goal is to describe various methods in which an analyst can share their data with minimal friction. We steer clear of idealitstic ideas such as the FAIR data principles [@Wilkinson2016] since they still do not help a researcher share their data. We also skip the discussion around citation and credit because data citations are still poorly tracked and there is no standardized metric or a h-index equivalent for data as of this writing.

@@

The first two of these challenges have largely been solved (REFERENCE).

@@

Providing the right amount of documentation for your data is challenging. Too much is overwhelming; too little and you don't know anything. To illustrate this, imagine two datasets. The first is contained in a folder with dozens of data files and several papers on the topic. The second contains only a single CSV file. The first is too much; understanding it requires a high time cost. The second is too little; we don't know anything about what the data is or why it was collected. The challenge is to provide the right level of detail, somewhere between these two extremes.

@@

One low cost and easy way to distribute data alongside compute would be to package the datasets as a data only package or as part of something larger where the methods are being implemented. The R ecosystem has many noteworthy examples of data-only packages.  One exemplar is the `nycflights13` package by Hadley Wickham (https://cran.r-project.org/web/packages/nycflights13/index.html). This package makes available airline data for all flights departing New York city in 2013 in a tidy format, with distinct tables for metadata on airlines, airports, weather, and planes. The package not only provides ready to use binary data but also ships raw data (in a folder called `data-raw`) along with scripts used to clean them. The package was originally created as a way to provide example data to teach tidy data principles and serves as a great model for how one can publish a data package in R.

@@

Of the 15539 packages on Central R Archive Network (CRAN), 6278 contain
datasets either as binary data (5903 packages) or as external datasets
(766). Binary files comprise a bulk of the data shipped in the `data`
folder (68.06%) with other plain text formats such as txt, CSV, dat,
`json` comprising less than one percent of data shipped in packages.

@@

Another common situation that researchers face is in dealing with data that fall somewhere between small (tabular data such as CSV that are in the order of bytes to megabytes to gigabytes that can be easily compressed) to large (what falls under the umbrella of `big data`). The happy medium is generally data that are too big to fit on memory of most standard machines, but can successfully fit on disk (https://ropensci.github.io/arkdb/articles/articles/noapi.html). In this case, users who do not have the support to maintain resource intensive database servers can instead rely on light serverless databases such as MonetDB or SQLite. These databases provide disk based storage and using language agnostic interfaces, a analyst can easily query these data in manageable chunks that don't overrun memory limitations. Using software packages such as `arkdb` (https://ropensci.github.io/arkdb/index.html) one could easily chunk large data from flat text files to these lite databases without running into memory limitations.

@@

In **Domain Specific Venue**, data can be published in journal data papers, or venues specific to the type of data. For example, in astronomy the data is hosted at the SDSS (SDSS), Genetic data can be hosted at GenBank (@GenBank), and so on.  Some other places to put data include the open science framework: https://osf.io/, Dataverse, https://dataverse.org/, and Data World, https://data.world/. Other datasets might be produced

The purpose, use, and origin of the data is an important component to consider. Data for research has a different domain compared to data collected by governments, or by journalists. Many governments or civil organizations are now making their own data available through a government website interface. Media and journalism groups are also making their data available either through places like GitHub (https://github.com/the-pudding/data), or organizations such as Data World, or may self host their data (https://data.fivethirtyeight.com/).

@@

Below is an example data dictionary table taken from the [Tidy Tuesday repository on incarceration trends](https://github.com/rfordatascience/tidytuesday). This includes information on the variable the class (type) of the variable, and a longer description of the variable.

@@

Other licenses or notices to be aware of are **copyrighted data**, and **data embargoes**. If you are working with data **already copyrighted**, (for example under CC BY or CC0), you must give follow appropriate guidelines for giving credit. Data may also be under **embargo**. This means data cannot be shared more widely until a specific release time. If sharing data under an embargo, include detailed information on the embargo requirements in: the README, and in separate correspondence with those who receive the data.

@@

TODO: Needs a succinct summary here

@@

4. **Provide a machine readable format** for your data that describes what each column contains. This can be a json schema file, which also helps websites like Google Dataset Search automatically discover your data.
5. **Provide the data in its rawest form** in a folder called "data-raw". This
 ensures provenance of your data by keeping a master copy from which all
other tidied up data can be created.
6. **Provide [open source?] scripts** used to clean data from rawest form into tidy/analysis ready format. These describe the steps taken to prepare the data,
which helps explain and document any decisions taken during the cleaning phase.
These should ideally operate on some raw data stored in the "data-raw" folder (rule 5).
7. **Keep additional copies in more accessible locations**: Even if you archive the data somewhere long-term, keep a second (or additional copies) in locations that are more readily and easily accessible for data analysis workflows. These include services like GitHub, GitHub LFS and others where fast CDNs make access and reuse practical, but fall back to Zenodo or similar in case these services were to go away.
8. **Use a hash function like MD5 checksum** to ensure that the data you deposited are the same that are being made available to others. Hash functions not only remove redundancy by removing the need to download the same data again, but also help ensure that the data are valid.
9. **Only add a citation if your data has a DOI**. A citation only makes sense when your data has a DataCite compliant DOI, which is automatically provided when data is published in repositories like Zenodo and Dryad. Even then a citation may not accrue references, but without a DOI, it is guaranteed not to.
10. Store a copy of your data in simple formats such as CSV: Plain text CSV files follow a flat and extremely simple schema and can be opened by any text editor. These files can also be easily parsed by any data science programming language. Modern software packages allow for fast reads without consuming too much memory and these tools also allow for easy modification of the data.
